---
title: "Inspiration_from_class"
output: html_document
---

# Instructions

* Compute the municipalities centroids with `st_centroid()` and store as an object named `centroids`. Only one argument is required.
* Plot the geometry of `municipalities` object with the centroids on top. This requires two calls to `plot()` run together. The second will need `add = TRUE`.
* Can you vary the size of centroids by the number of Viking finds contained by the municipality/or by their density?

```{r mun-centroids, eval = FALSE}
centroids <- st_centroid(mun_25832)

mapview(mun_25832, col.regions = "beige", legend = FALSE) + 
  mapview(centroids,cex = "v_density", zcol = "viking")

mapview(mun_25832, col.regions = "beige", legend = FALSE) + 
  mapview(centroids,cex = "monuments" , zcol = "viking" )
```

# Task 4: Computing bounding boxes

You can compute bounding boxes around vector data using `sf`. These can help you, for example, create polygons to clip layers to a common area for an analysis (e.g to limit area to Aarhus municipality) or identify regions of influence.

In the `sf` package, there is a function for extracting the bounding box coordinates, if that's all you need, this is `st_bbox()`. More likely you'll want to create a new `sf` object (a polygon) from those coordinates and to do this `sf` provides the `st_make_grid()` function.

`st_make_grid()` can be used to make a multi-row and multi-column grid covering your input data but it can also be used to make a grid of just one cell (a bounding box). To do this, you need to specify the number of grid cells as `n = 1`.

## Instructions

* Filter out the Aarhus (or another) kommune from the `mun_25832` sf object and create a new `aarhus_mun` object. How exactly is Aarhus spelled and which column do you use? 
* Subset the `viking` file to Aarhus kommune and plot these `a_viking` points over the `aarhus_mun` object - they should overlap.
* Compute the bounding box coordinates with `st_bbox()`. What is the result?
* Create a single polygon bounding box around the Aarhus municipality using `st_make_grid()` with an argument of `n = 1`.
* Create a plot of the `municipalities`, highlighting the Aarhus municipality and `aarhus_box` (just the geometry), the new box on top of it with `a_viking` points on top.

```{r Ar-box, eval= FALSE}
# Filter Aarhus [or another municipality] out of the municipalities 
unique(mun_25832$NAME_2)
aarhus_mun <- mun_25832 %>% 
  filter(NAME_2 == "Århus")

# Filter out also the viking finds in Aarhus
a_viking <- viking %>% 
  filter(NAME_2 == "Århus")

# Plot the Aarhus municipality and the local viking finds
plot(st_geometry(aarhus_mun), col = "grey", border = "white")
plot(st_geometry(a_viking), add = TRUE, pch = 16, col = "forestgreen")

# Check the coordinates of the Aarhus bounding box
st_bbox(aarhus_mun)

# Compute a bounding box polygon
aarhus_box <- st_make_grid(aarhus_mun, n=1)

# Plot all the DK municipalities, add the shelters and the new box outlining the Aarhus municipality
plot(st_geometry(mun_25832), col = "grey", border = "white")
plot(a_viking$geometry, pch = 16, col = "forestgreen", add = T)
plot(aarhus_box, lwd = 2, add = T)
```
Bounding box completed! You can see in the plot that the bounding box is the minimum rectangular polygon that fits all the monuments. You might now clip a series of layers to this box for a consistent extent over the Aarhus municipality. 


Let's build on what we learned and add one more geoprocessing task: computing buffers- which are the bread-and-butter of spatial analysis. The resulting buffers have a wide range of uses like, for example, identifying the number of roads within one kilometer of a school or computing the number of hazardous waste sites near sensitive natural areas.


* Use `st_intersection()` to select shelters that fall inside your district's bounding box (you can also create a another buffer around your district if that makes more sense).

```{r Ar-box, eval= FALSE}
# Load districts
districts <- read_sf("../data/mystery/districts2023.geojson")

mapview(districts, zcol = "prog_distrikt_navn", alpha = 0.5)

# Create bounding box around district
bb <- districts %>% 
  filter(prog_distrikt_navn == "Christiansbjerg") %>% 
  st_bbox() %>% 
  st_make_grid(n =1)
mapview(bb) + mapview(shelters)

# Load shelters
shelters <- read_sf("../data/sheltersAarhus.geojson")

# Create your address as a dataframe with lat long data, following this pattern:
df <- data.frame(place = c("Aarhus Domkirke", 
                           "Christianskirken",
                           "Nobelparken"),
                 longitude=c(10.2102364,10.1949197,10.2033485),
                 latitude=c(56.1574919,56.1772228,56.174346))

df <- df %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Compute a buffer around your address

df_buf <- st_buffer(df, dist=300)
?st_buffer()
mapview(df_buf)


mapview(shelters) + mapview(df_buf)
# Calculate which shelters fall within the buffer or bounding box of your place
my_shelters <- st_intersection(shelters, df_buf)
mapview(my_shelters) + mapview(df_buf)

# Visualize your buffer and bounding box with intersecting shelters
my_shelters2 <- st_intersection(shelters, bb)
mapview(bb) + mapview(my_shelters2)
```


# Map Overlay with Isochrones

We can visualize this situation in central Jutland with layered isochrones. We already used this technique to show multiple drive times around the Aarhus University earlier in this tutorial. In this case, we will use `mb_isochrone()` to generate dozens of isochrones, then visualize them simultaneously to illustrate an accessibility gradient in the region.


To generate the accessibility surface raster, we will need to apply a *coordinate system transformation* to “project” our data to two-dimensional coordinates. This will allow us to specify the raster’s resolution in meters. We generate a 100m resolution raster, and use the `rasterize()` function in `terra` to allocate the minimum overlapping value from our isochrones to each grid cell. The result can then be mapped with Leaflet’s `addRasterImage()` function.


# Create a RasterLayer (SpatRaster in terra) with the specified resolution of 100m
template <- rast(isos_32, resolution = 100)

library(terra)
# Use the rasterize function from terra, specifying the field and aggregation function
iso_surface <- rasterize(isos_32, template, field = "time", fun = min, background=NA)

# Color mapping setup using terra (you may need to install 'viridis' for color palettes)
library(viridis)
cols <- viridis(100, option = "D")

# Plotting - note that terra uses plot() as the function of choice
plot(iso_surface, col=cols, main="Drive-time to Skejby Hospital")



Measuring distance between feature sets is a component of spatial analysis 101 -- a core skill for any analyst. There are several functions in `base` R as well as in the packages `rgeos` and `geosphere` to compute distances, but the `st_distance()` function from `sf` provides a useful feature-to-feature distance matrix as output and can be used for most distance calculation needs, including Euclidean and geodesic distances.




# 3. Construct a routable street network
In this step, we convert the street centerlines into a routable network structure. This basically means that we make it explicit which street segments share an endpoint, such that we know how streets are connected, and hence, how someone can travel from A to B. This can be done very easily with the `sfnetworks` package. Note that for now we simply assume each street can be traveled in both directions, and the network is therefore undirected.

```{r create_routable_network}
network <-sfnetworks::as_sfnetwork(streets, directed = FALSE)

par(mar = c(0, 0, 0, 0))
plot(network, cex = 0.5)
```

However, streets in OpenStreetMap are not always mapped with a network structure in mind. It happens often that connected street segments cross each other without sharing an endpoint. This results in many disconnected sets of streets in the network, as shown below. Even our small network has over 5000 disconnected components, with the largest of them containing around 20% of all nodes.

```{r count_network_components}
tidygraph::with_graph(network, tidygraph::graph_component_count())
```

```{r analyze_network_connectivity}
node_count_all <-network %>%
  tidygraph::with_graph(tidygraph::graph_order())

node_count_largest <-network %>%
  tidygraph::convert(tidygraph::to_largest_component) %>%
  tidygraph::with_graph(tidygraph::graph_order())

round(node_count_largest / node_count_all * 100)
```


Luckily, the `sfnetworks` package contains functions to pre-process and clean networks. One of them is `to_spatial_subdivision()`, which connects streets when they share internal vertices rather than only endpoints. This results in a network that still has some disconnected components (which is to be expected since we cut the network at the buffer boundaries), but with the largest of them containing almost all nodes.

There are many more network pre-processing and cleaning functions in `sfnetworks`. See the `sfnetworks documentation` for details.
```{r preprocess_network_connectivity}
network <-network %>%
  tidygraph::convert(sfnetworks::to_spatial_subdivision)

par(mar = c(0, 0, 0, 0))
plot(network, cex = 0.5)

tidygraph::with_graph(network, tidygraph::graph_component_count())

node_count_all <-network %>%
  tidygraph::with_graph(tidygraph::graph_order())

node_count_largest <-network %>%
  tidygraph::convert(tidygraph::to_largest_component) %>%
  tidygraph::with_graph(tidygraph::graph_order())

round(node_count_largest / node_count_all * 100)
```


Now we can select the largest component without having to worry we loose a lot of data.
```{r select_largest_network_component}
network <-network %>%
  tidygraph::convert(tidygraph::to_largest_component)

par(mar = c(0, 0, 0, 0))
plot(network, cex = 0.5)
```


```{r}
library(sfnetworks)
library(dplyr)
library(tidygraph)
library(sf)

valid_network <- valid_network %>%
  activate("edges") %>%
  mutate(weight = edge_length())


# shortest path
path <- st_network_paths(
  valid_network,
  from = venue,         # den gamle by
  to = access[1, ],     # shelter
  weights = "time"
)

# Extract and plot the route
route_edges <- valid_network %>%
  activate("edges") %>%
  slice(path$edge_paths[[1]]) %>%
  st_as_sf()

# Inspect the route
  mapview::mapview(route_edges, color = "red", lwd = 4) +
  mapview::mapview(venue, col.region = "blue", cex = 8) +
  mapview::mapview(access[1, ], col.region = "green", cex = 8)
```
