---
title: 'CDS Spatial Analytics Exam: Pathway Comparison'
author: "Zofia Radwańska"
date: "2026-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#loading in the packages
```{r}
install.packages("Cairo")
install.packages("ragg")
library(pacman)
pacman::p_load(tidyverse,
               readr,
               dplyr,
               ggplot2,
               sf,
               tidyr,
               ggdendro,
               stats,
               ggrastr,
               patchwork)

# not using CRS
sf::sf_use_s2(FALSE)
drop_crs <- function(x) {
  st_crs(x) <- NA
  x
}
```

# LOADING DATA
```{r}
# loading trajectory data and drop CRS
geo_trajectories <- st_read("data/all_trajectories_with_floor.geojson")
geo_trajectories$floorNumber <- as.integer(geo_trajectories$floorNumber)
geo_trajectories <- drop_crs(geo_trajectories)


#floor plan + drop crs
floor_0_plan_sf <- st_read("data/geojson_converted_objects/floor_0_all_objects.geojson")
floor_1_plan_sf <- st_read("data/geojson_converted_objects/floor_1_all_objects.geojson")
floor_2_plan_sf <- st_read("data/geojson_converted_objects/floor_2_all_objects.geojson")
floor_0_plan_sf <- drop_crs(floor_0_plan_sf)
floor_1_plan_sf <- drop_crs(floor_1_plan_sf)
floor_2_plan_sf <- drop_crs(floor_2_plan_sf)

#room plan - polygons and drop CRS
floor_0_rooms <- st_read("data/polygons/polygons_floor_0.geojson")
floor_1_rooms <- st_read("data/polygons/polygons_floor_1.geojson")
floor_2_rooms <- st_read("data/polygons/polygons_floor_2.geojson")
floor_0_rooms <- drop_crs(floor_0_rooms)
floor_1_rooms <- drop_crs(floor_1_rooms)
floor_2_rooms <- drop_crs(floor_2_rooms)

#renaming columns
floor_0_rooms <- floor_0_rooms %>% select(room_id, geometry)
floor_1_rooms <- floor_1_rooms %>% mutate(room_id = floor_plan)
floor_1_rooms <- floor_1_rooms %>% select(room_id, geometry)
floor_2_rooms <- floor_2_rooms %>% mutate(room_id = fid)
floor_2_rooms <- floor_2_rooms %>% select(room_id, geometry)

plot(st_geometry(floor_0_plan_sf), main = "Museum Floor Plan - Ground Floor", col = "pink")
```

# PREPROCESSING
## split trajectories by floor
```{r}
#trajectories based on floor
trajectory_f00 <- geo_trajectories %>% 
  dplyr::filter(.data$floorNumber == 0)
trajectory_f01 <- geo_trajectories %>% 
  dplyr::filter(.data$floorNumber == 1)
trajectory_f02 <- geo_trajectories %>% 
  dplyr::filter(.data$floorNumber == 2)

#check how many points per floor
cat("Floor 0:", nrow(trajectory_f00), "points\n")
cat("Floor 1:", nrow(trajectory_f01), "points\n")
cat("Floor 2:", nrow(trajectory_f02), "points\n")
```

## filter trajectory by room polygons
```{r}
# filter out trajectory points that fall outside of the room polygons
filter_trajectories_by_rooms <- function(trajectory_data, room_polygons) {
  
  # find which trajectory points are inside any room polygon
  points_in_rooms <- st_intersects(trajectory_data, 
                                   st_union(room_polygons), 
                                   sparse = FALSE)
  
  # filter trajectory data to keep only points inside rooms
  filtered_trajectories <- trajectory_data[points_in_rooms, ]
  
  cat("Original points:", nrow(trajectory_data), "\n")
  cat("Points inside rooms:", nrow(filtered_trajectories), "\n")
  cat("Points removed:", nrow(trajectory_data) - nrow(filtered_trajectories), "\n")
  cat("Retention rate:", 
      round(100 * nrow(filtered_trajectories) / nrow(trajectory_data), 2), "%\n")
  
  return(filtered_trajectories)
}

trajectory_f0_filtered <- filter_trajectories_by_rooms(trajectory_f00, floor_0_rooms)
trajectory_f1_filtered <- filter_trajectories_by_rooms(trajectory_f01, floor_1_rooms)
trajectory_f2_filtered <- filter_trajectories_by_rooms(trajectory_f02, floor_2_rooms)
```

## trajectory extraction and conversion
```{r}
# extract and convert trajectories to LINESTRING per participant per floor
extract_trajectories_as_lines <- function(trajectory_data, floor_number) {
  
  # filter by floor and sort by participant and timestamp
  floor_data <- trajectory_data %>%
    filter(floorNumber == floor_number) %>%
    arrange(id, timestamp)
  
  # get unique participant IDs 
  participant_ids <- unique(floor_data$id)
  
  # create linestring for each participant
  trajectories_list <- vector("list", length = length(participant_ids))
  names(trajectories_list) <- as.character(participant_ids)  # use actual ids
  
  for (i in seq_along(participant_ids)) {
    pid <- participant_ids[i]
    
    # filter to single participant
    pdata <- floor_data %>% filter(id == pid)
    
    # extract coordinates
    coords <- st_coordinates(pdata$geometry)
    
    # create linestring (at least 2 points)
    if (nrow(coords) >= 2) {
      trajectories_list[[as.character(pid)]] <- st_linestring(coords)
    } else {
      trajectories_list[[as.character(pid)]] <- NULL
    }
  }
  
  # remove null entries
  trajectories_list <- trajectories_list[!sapply(trajectories_list, is.null)]
  
  return(trajectories_list)
}

```

# ANALYSIS FUNCTIONS
## Fréchet distance calculation
```{r}
# calculate pairwise Fréchet distance matrix for trajectories
calculate_frechet_matrix <- function(trajectories) {
  
  n <- length(trajectories)
  traj_names <- names(trajectories)
  
  # convert list to sf collection for st_distance
  traj_sfc <- st_sfc(trajectories)
  
  # calculate pairwise distances
  dist_matrix <- st_distance(traj_sfc, which = "Frechet")
  
  # set rownames and colnames from original trajectory names
  rownames(dist_matrix) <- traj_names
  colnames(dist_matrix) <- traj_names
  
  return(as.matrix(dist_matrix))
}
```

## variability metrics
```{r}
# calculate trajectory variability metrics per floor
calculate_variability_metrics <- function(dist_matrix) {
  
  # remove diagonal (self-distances = 0)
  upper_tri <- dist_matrix[upper.tri(dist_matrix)]
  
  metrics <- list(
    mean_distance = mean(upper_tri),
    median_distance = median(upper_tri),
    sd_distance = sd(upper_tri),
    min_distance = min(upper_tri),
    max_distance = max(upper_tri),
    cv_distance = sd(upper_tri) / mean(upper_tri),  # coefficient of variation
    q25 = quantile(upper_tri, 0.25),
    q75 = quantile(upper_tri, 0.75),
    iqr = IQR(upper_tri),
    n_trajectories = nrow(dist_matrix),
    all_distances = upper_tri
  )
  
  return(metrics)
}

# calculate distance from mean trajectory (centroid path)
calculate_distance_from_centroid <- function(trajectories, dist_matrix) {
  
  # distance from centroid = mean distance to all other trajectories
  distance_from_centroid <- rowMeans(dist_matrix)
  
  return(distance_from_centroid)
}
```

# VISUALISATION FUNCTIONS
## colour pallett
```{r}
palette_loud <- list(
  floor0 = c(dark = "#92113AFF", base = "#C1395EFF"),
  floor1 = c(dark = "#DFA90DFF", base = "#FFD84AFF"),
  floor2 = c(dark = "#5F7F00FF", base = "#AEC17BFF")
)

# get plot colour helper function
get_floor_colors <- function(floor_name, palette_loud) {
  # Handle different floor name formats
  # "Floor 0" or "floor 0" or "floor_0" -> "floor0"
  floor_key <- tolower(gsub("\\s+|_", "", floor_name))
  if (!floor_key %in% names(palette_loud)) {
    stop(paste("Unknown floor:", floor_name))
  }
  palette_loud[[floor_key]]
}

```

## heatmap of pathway distnaces (similarity)
```{r}
# plot frechet distance heatmap with dendrogram
plot_distance_heatmap <- function(dist_matrix, clusters = NULL, floor_name = "Floor 0", 
                                  title_prefix = "Fréchet Distance Matrix") {
  
  #convert to regular matrix
  if (inherits(dist_matrix, "dist")) {
    dist_matrix <- as.matrix(dist_matrix)
  } else if (inherits(dist_matrix, "units")) {
    dist_matrix <- as.numeric(dist_matrix)
  } else {
    dist_matrix <- as.matrix(dist_matrix)
  }
  
  # ensure col and row names work
  if (is.null(rownames(dist_matrix))) {
    rownames(dist_matrix) <- 1:nrow(dist_matrix)
  }
  if (is.null(colnames(dist_matrix))) {
    colnames(dist_matrix) <- 1:ncol(dist_matrix)
  }
  
  # perform clustering if not provided
  hc <- hclust(as.dist(dist_matrix), method = "ward.D2")
  order_idx <- hc$order
  
  # reorder matrix by clustering
  dist_ordered <- dist_matrix[order_idx, order_idx]
  
  # convert to long format for ggplot
  heatmap_data <- as.data.frame(dist_ordered) %>%
    rownames_to_column("Trajectory1") %>%
    pivot_longer(-Trajectory1, names_to = "Trajectory2", values_to = "Distance") %>%
    mutate(
      Distance = as.numeric(Distance),
      Trajectory1 = factor(Trajectory1, levels = rownames(dist_ordered)),
      Trajectory2 = factor(Trajectory2, levels = colnames(dist_ordered))
    )
  
  # get floor colors
  floor_colors <- get_floor_colors(floor_name, palette_loud)
  
  # create heatmap
  p <- ggplot(heatmap_data, aes(x = Trajectory2, y = Trajectory1, fill = Distance)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "white",
    mid = "grey90",
    high = unname(floor_colors["base"]),
    midpoint = median(heatmap_data$Distance, na.rm = TRUE)
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 7),
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  ) +
  ggtitle(paste0(title_prefix, " - ", floor_name))
  
  return(p)
}
```

## trajectory density heatmap (over the floorplan)
```{r}
# plot trajectory density heatmap on floor
plot_trajectory_density <- function(trajectory_data, floor_plan, floor_name = "Floor 0", bins = 30) {
  
  # extract coordinates
  coords <- st_coordinates(st_cast(trajectory_data, "POINT"))
  
  # create plot
  floor_colors <- get_floor_colors(floor_name, palette_loud)

  p <- ggplot() +
    geom_sf(data = floor_plan, fill = "white", color = "black", size = 0.5) +
  
    geom_bin2d(
      data = as.data.frame(coords),
      aes(x = X, y = Y),
      bins = bins,
      alpha = 0.7
    ) +
  
    scale_fill_gradient(
      low = "white",
      high = unname(floor_colors["base"]),
      name = "Point\nDensity"
    ) +
    
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.text = element_blank(),
      axis.ticks = element_blank()
    ) +
    labs(title = paste0("Movement Density - ", floor_name))
    
  return(p)
}
```

## trajectory density scatter plot (over the floorplan)
```{r}
plot_trajectory_density_scatter <- function(trajectory_data, floor_plan, 
                                           floor_name = "Floor 0") {
  
  floor_colors <- get_floor_colors(floor_name, palette_loud)
  
  # create plot
  p <- ggplot() +
    geom_sf(data = floor_plan, fill = "white", color = "black", size = 0.5) +
    geom_sf(
      data = trajectory_data,
      color = unname(floor_colors["base"]),
      alpha = 0.15,
      size = 0.8
    ) +
    coord_sf(expand = FALSE) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold", margin = margin(b = 10)),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      panel.background = element_blank(),
      plot.margin = margin(0, 0, 0, 0)
    ) +
    labs(title = paste0("Movement Density (Point Scatter) - ", floor_name),
         caption = "Darker areas = more overlapping points")
  
  return(p)
}
```

## density distributions
```{r}
plot_distance_distributions <- function(metrics_list) {
  
  comparison_data <- data.frame(
    floor = rep(names(metrics_list), 
                times = sapply(metrics_list, function(x) length(x$all_distances))),
    distance = unlist(lapply(metrics_list, function(x) x$all_distances))
  )
  
  # Convert floor_0 format to Floor 0 for display
  comparison_data$floor_display <- sapply(comparison_data$floor, function(f) {
    paste0("Floor ", substr(f, nchar(f), nchar(f)))
  })
  
  floor_fill_palette <- c(
    floor_0 = unname(palette_loud$floor0["base"]),
    floor_1 = unname(palette_loud$floor1["base"]),
    floor_2 = unname(palette_loud$floor2["base"])
  )
  
  ggplot(comparison_data, aes(x = distance, fill = floor)) +
    geom_histogram(alpha = 0.5, bins = 30, position = "identity") +
    geom_density(alpha = 0.3, linewidth = 1) +
    
    facet_wrap(~floor_display, ncol = 3) +
    
    scale_fill_manual(values = floor_fill_palette) +
    
    theme_minimal() +
    theme(
      legend.position = "none",
      strip.text = element_text(face = "bold")
    ) +
    labs(
      #title = "Distribution of Trajectory Distances per Floor",
      x = "Fréchet Distance",
      y = "Frequency"
    )
}
```

## cv comparison
```{r}
plot_cv_comparison <- function(metrics_list) {
  cvs <- sapply(metrics_list, function(x) x$cv_distance)
  n_vals <- sapply(metrics_list, function(x) length(x$all_distances))
  
  # 95% CI for CV (using delta method)
  cv_ci_lower <- cvs - 1.96 * cvs / sqrt(2 * n_vals)
  cv_ci_upper <- cvs + 1.96 * cvs / sqrt(2 * n_vals)
  
  cv_df <- data.frame(
    floor = names(metrics_list),
    cv = cvs,
    ci_lower = cv_ci_lower,
    ci_upper = cv_ci_upper
  )
  
  # Map floor names to palette keys (floor_0 -> floor0, floor_1 -> floor1, etc.)
  palette_map <- list(
    floor_0 = palette_loud$floor0["base"],
    floor_1 = palette_loud$floor1["base"],
    floor_2 = palette_loud$floor2["base"]
  )
  
  fill_values <- setNames(
    sapply(cv_df$floor, function(f) unname(palette_map[[f]])),
    cv_df$floor
  )
  
  ggplot(cv_df, aes(x = floor, y = cv, fill = floor)) +
    geom_col(alpha = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    
    scale_fill_manual(values = fill_values) +
    
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
    ) +
    labs(
      title = "Coefficient of Variation Across Floors",
      subtitle = "With 95% confidence intervals",
      x = "Floor",
      y = "Coefficient of Variation",
      caption = "Lower CV = more uniform trajectories"
    )
}
```

## cv bootstrap distributions
```{r}
plot_cv_bootstrap_distributions <- function(metrics_list) {
  floor_names <- names(metrics_list)
  distances_list <- lapply(metrics_list, function(x) x$all_distances)
  
  set.seed(42)
  n_boots <- 5000
  boot_data <- data.frame()
  
  for (i in 1:(length(floor_names) - 1)) {
    for (j in (i + 1):length(floor_names)) {
      d1 <- distances_list[[i]]
      d2 <- distances_list[[j]]
      
      cv_diffs <- numeric(n_boots)
      for (b in 1:n_boots) {
        d1_boot <- sample(d1, replace = TRUE)
        d2_boot <- sample(d2, replace = TRUE)
        
        cv1_boot <- sd(d1_boot) / mean(d1_boot)
        cv2_boot <- sd(d2_boot) / mean(d2_boot)
        
        cv_diffs[b] <- cv1_boot - cv2_boot
      }
      
      # Convert floor_0 format to Floor 0 for display
      floor_i_display <- paste0("Floor ", substr(floor_names[i], nchar(floor_names[i]), nchar(floor_names[i])))
      floor_j_display <- paste0("Floor ", substr(floor_names[j], nchar(floor_names[j]), nchar(floor_names[j])))
      comparison_label <- paste0(floor_i_display, " vs ", floor_j_display)
      
      boot_data <- rbind(boot_data, data.frame(
        comparison = comparison_label,
        cv_diff = cv_diffs
      ))
    }
  }
  
  # Create palette with matching comparison labels
  palette_loud <- list(
    floor0 = c(dark = "#92113AFF", base = "#C1395EFF"),
    floor1 = c(dark = "#DFA90DFF", base = "#FFD84AFF"),
    floor2 = c(dark = "#5F7F00FF", base = "#AEC17BFF")
  )
  
  boot_palette <- c(
    "Floor 0 vs Floor 1" = unname(palette_loud$floor0["base"]),
    "Floor 0 vs Floor 2" = unname(palette_loud$floor0["base"]),
    "Floor 1 vs Floor 2" = unname(palette_loud$floor1["base"])
  )
  
  ggplot(boot_data, aes(x = cv_diff, fill = comparison)) +
    geom_density(alpha = 0.6) +
    geom_vline(xintercept = 0, linetype = "dashed", size = 1, color = "red") +
    facet_wrap(~comparison) +
    scale_fill_manual(values = boot_palette) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
    ) +
    labs(
      title = "Bootstrap Distribution of CV Differences",
      x = "CV Difference",
      y = "Density",
      caption = "Red line = null hypothesis (no difference)"
    )
}
```

# STATISTICAL TESTING - CV COMPARISON
Bootstrap method for comparison of coefficients of variability.
```{r}
compare_cv_statistically <- function(metrics_list) {
  
  floor_names <- names(metrics_list)
  n_floors <- length(metrics_list)
  cvs <- sapply(metrics_list, function(x) x$cv_distance)
  
  cat("=== COEFFICIENT OF VARIATION COMPARISON ===\n\n")
  
  cv_summary <- data.frame(
    floor = floor_names,
    cv = round(cvs, 4),
    n_comparisons = sapply(metrics_list, function(x) length(x$all_distances))
  )
  print(cv_summary)
  
  # bootstrap test
  cat("\n=== BOOTSTRAP TEST (non-parametric) ===\n")
  cat("Resamples data to test if CV differences are significant\n\n")
  
  distances_list <- lapply(metrics_list, function(x) x$all_distances)
  set.seed(42)
  n_boots <- 10000
  boot_results <- list()
  
  for (i in 1:(n_floors - 1)) {
    for (j in (i + 1):n_floors) {
      test_name <- paste0(floor_names[i], "_vs_", floor_names[j])
      
      d1 <- distances_list[[i]]
      d2 <- distances_list[[j]]
      cv_diff_obs <- cvs[i] - cvs[j]
      
      # bootstrap
      cv_diffs <- numeric(n_boots)
      for (b in 1:n_boots) {
        d1_boot <- sample(d1, replace = TRUE)
        d2_boot <- sample(d2, replace = TRUE)
        
        cv1_boot <- sd(d1_boot) / mean(d1_boot)
        cv2_boot <- sd(d2_boot) / mean(d2_boot)
        
        cv_diffs[b] <- cv1_boot - cv2_boot
      }
      
      # two-tailed p-value
      p_value_boot <- mean(abs(cv_diffs) >= abs(cv_diff_obs))
      
      # 95% CI
      ci_lower <- quantile(cv_diffs, 0.025)
      ci_upper <- quantile(cv_diffs, 0.975)
      
      boot_results[[test_name]] <- list(
        cv_diff_observed = cv_diff_obs,
        p_value = p_value_boot,
        ci_lower = ci_lower,
        ci_upper = ci_upper
      )
      
      cat(test_name, "\n")
      cat("  CV difference:", round(cv_diff_obs, 4), "\n")
      cat("  Bootstrap p-value:", round(p_value_boot, 4), 
          ifelse(p_value_boot < 0.001, "***", 
                 ifelse(p_value_boot < 0.01, "**", 
                        ifelse(p_value_boot < 0.05, "*", "ns"))), "\n")
      cat("  95% CI:", round(ci_lower, 4), "to", round(ci_upper, 4), "\n\n")
    }
  }
  
  result <- list(
    cv_summary = cv_summary,
    bootstrap_tests = boot_results
  )
  
  return(result)
}
```

# MAIN ANALYSIS PIPELINE
```{r}
# extract trajectories per floor as linestrings
trajectories_floor_0 <- extract_trajectories_as_lines(trajectory_f0_filtered, floor_number = 0)
trajectories_floor_1 <- extract_trajectories_as_lines(trajectory_f1_filtered, floor_number = 1)
trajectories_floor_2 <- extract_trajectories_as_lines(trajectory_f2_filtered, floor_number = 2)

# calculate frechet distance matrices
dist_matrix_f0 <- calculate_frechet_matrix(trajectories_floor_0)
dist_matrix_f1 <- calculate_frechet_matrix(trajectories_floor_1)
dist_matrix_f2 <- calculate_frechet_matrix(trajectories_floor_2)

# calculate variability metrics
metrics_f0 <- calculate_variability_metrics(dist_matrix_f0)
metrics_f1 <- calculate_variability_metrics(dist_matrix_f1)
metrics_f2 <- calculate_variability_metrics(dist_matrix_f2)

metrics_list <- list(floor_0 = metrics_f0, floor_1 = metrics_f1, floor_2 = metrics_f2)
```

# RESULTS: Statistical testing
```{r}
cv_stats <- compare_cv_statistically(metrics_list)
```

# RESULTS - VISUALISATION
```{r}
# trajectory similarity
p_heatmap_f0 <- plot_distance_heatmap(dist_matrix_f0, floor_name = "Floor 0")
p_heatmap_f1 <- plot_distance_heatmap(dist_matrix_f1, floor_name = "Floor 1")
p_heatmap_f2 <- plot_distance_heatmap(dist_matrix_f2, floor_name = "Floor 2")

# trajectory density
p_density_f0 <- plot_trajectory_density(trajectory_f0_filtered, floor_0_plan_sf, "Floor 0")
p_density_f1 <- plot_trajectory_density(trajectory_f1_filtered, floor_1_plan_sf, "Floor 1")
p_density_f2 <- plot_trajectory_density(trajectory_f2_filtered, floor_2_plan_sf, "Floor 2")

# trajectory density (scatterplot)
p_density_scatter_f0 <- plot_trajectory_density_scatter(trajectory_f0_filtered, floor_0_plan_sf, "Floor 0")
p_density_scatter_f1 <- plot_trajectory_density_scatter(trajectory_f1_filtered, floor_1_plan_sf, "Floor 1")
p_density_scatter_f2 <- plot_trajectory_density_scatter(trajectory_f2_filtered, floor_2_plan_sf, "Floor 2")

# distance distirbution
p_distributions <- plot_distance_distributions(metrics_list)

# cv comparison
p_cv <- plot_cv_comparison(metrics_list)
p_cv_boot <- plot_cv_bootstrap_distributions(metrics_list)

# combined scatter density plot
combined_plot <- (p_density_scatter_f0 | p_density_scatter_f1 | p_density_scatter_f2) +
  plot_layout(
    ncol = 3,
    widths = c(1, 1, 1)
  ) +
  plot_annotation(
    title = "Movement Density (Point Scatter)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.margin = margin(0, 0, 10, 0)
    )
  ) &
  theme(
    plot.margin = margin(0, 2, 0, 2)
  )

combined_plot

combined_plot

combined_plot

